Separating random and deterministic sources of computational noise in explore-exploit decisions

### 2024-05-13 manuscript/2024 - Plot Comp Bio/manuscript_DeterministicRandomNoise.tex
intro/methods

\title{\replaced[id=siyu]{Separating random and deterministic sources of computational noise in explore-exploit decisions}{The nature of decision noise in random exploration}}
\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

@@

	\newpage
	\begin{abstract}
Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. \added[id=bob]{A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random.} That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=bob]{By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random.} Using this approach, we find evidence that \added[id=siyu]{at least 15$\%$} of the variability in random exploration can be accounted for by deterministic processing of the stimulus. \added[id=bob]{Conversely, this suggests that up to 85$\%$ of the variability is truly `random,' although it is still possible that this variability is driven by deterministic factors not related to the stimulus.}
% This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. 
\added[id=bob]{Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.}

%suggests that at least some `random' exploration is not random, but leaves open the possibility that the 
%However, it is unclear whether this variability is truly random
%From a modeling perspective, behavioral variability is essentially the variance that can not be explained by a model and is modeled as the level of decision noise. However, what we have called "decision noise" in previous researches could actually just be missing deterministic components from the model, it is difficult to tell whether decision noise truly arises from a stochastic process. Here we show that, while both random and deterministic noise drive variability in behavior, the noise driving random exploration is predominantly random. This suggests that random exploration depends on adaptive noise processes in the brain which are subject to cognitive control.

\end{abstract}
	\newpage
	\section*{Introduction}

	%BOB SAYS: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT
	%SIYU Says: We can define "random noise" as "not a deterministic funciton of the stimulus" at the beginning, and use the term "random noise" in the results section. It feels a little lengthy to use "not a deterministic function of the stimulus" in the results section each time. I changed a few of them back to "random noise" in the results section, in places that I think random noise is better. See if you like the changes.

	%(BOB SAYS: TAKE OR LEAVE THIS CHANGE TO DATE, BUT SETTING UP AS A DATE ALLOWS YOU TO TALK ABOUT AVOID YOUR EX LATER ON, WHICH MAY BE A BETTER EXAMPLE OF A DETERMINISTIC PROCESS!).
	% Siyu Say: I kept this.

	Imagine trying to decide where to go to dinner on a date. You can go to your favorite restaurant, the one you both really enjoy and always go to, or you can try a new restaurant that you know nothing about. Such decisions, in which we must choose between a well-known `exploit' option and a lesser known `explore' option, are known as explore-exploit decisions.  From a theoretical perspective, making optimal explore-exploit choices, i.e. choices that maximize long-term reward, is computationally intractable in most cases \citep{eegittins74, eeBasu18}. In part because of this computational complexity, there is considerable interest in how humans and animals solve the explore-exploit dilemma in practice \added[id=siyu]{\citep{Mehlhorn15, SCHULZ20197,WILSON202149}}.
%\citep{eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90} \added[id=siyu]{, also known as random exploration}. In this strategy, \deleted[id=siyu]{the decision process between exploration and exploitation is corrupted by `decision noise', meaning that }high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. \added[id=siyu]{In modeling terms, random exploration works by adding `decision noise' to the value of the options such that sub-optimal exploratory options can sometimes have a higher total score (i.e., value + noise) than the exploit option and get chosen.} \replaced[id=siyu]{Such}{In theory, such} random exploration, is surprisingly effective and, if implemented correctly, can come close to optimal performance \added[id=siyu]{in some cases}\citep{eethompson33, eebridle90, eeAgrawal11, eeChapelle11}.

	It has recently been shown that humans appear to use random exploration and can increase  decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014} \added[id=bob]{ALSO CITE FINDLING ET AL.}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	\replaced[id=siyu]{One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' comes from a truly random process.}{One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random.} That is, whether behavioral variability is due to genuinely stochastic processes in the brain or whether it is due to deterministic processes that we \replaced[id=siyu]{failed to observe}{simply have not observed}. \added[id=bob]{Decision noise as defined in previous research essentially quantifies the extent to which behavior cannot be predicted by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise.} For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

	%because noise in previous work is defined as the component of the decision process that cannot be accounted for by a model. 
	%, 
	%For example, in the restaurant case above, if you happen to spot an old friend walking in to one of the restaurants, you may be more likely to follow them in. Such a process would be deterministic, in that it is driven by an observable stimulus, but unless a model of your behavior incorporated the presence of your friend,

	%. If the model did not consider that the agent is favoring the behavior of following a friend in a deterministic way, the behavior of going to a less favorable restaurant because of a friend will appear to be `random' when it is really a deterministic effect. Hence this deterministic factor will be modeled as random decision noise. Crucially, however, this deterministic source of variability is very much in the stimulus and if you saw the same friend go into the same restaurant at a later date you might follow them again. Conversely, truly `random' noise would arise from stochastic mental processes,

	%tossing a metaphorical coin in your head. Such a process would not be influenced by the friend going into the restaurant, and if you saw the same friend again, you might make a different choice.

	%A key limitation of our previous work was that the source of the decision noise used for exploration was left unknown.  In particular, due to the design of the task, we could not determine whether the adaptive decision noise that is linked to exploration arises deterministicly, in the input from the world, or is generated randomly, within the brain.

@@

    %* Difficulty with measuring random exploration, how do you know something is random, previous work model
    %* Model all the deterministic factor (contribute to noise)
    %* Have repeated games
    %* Missing deterministic dependent, much smaller than non-deterministic component
    %* Stochastic process…

	%In the restaurant case, an example of deterministic noise would be if you happen to spot an old friend walking in to one of the restaurants. Such an event is random, in that you couldn't predict it, and has the potential to alter your behavior as you're likely to follow your friend in. Crucially, however, this `randomness' is very much in the world and if you saw the same friend go into the same restaurant at a later date you might follow them  again.

	%Conversely, random noise would arise from stochastic neural processes tossing a metaphorical coin in your head. Such a process would not be influenced by the friend going into the restaurant, and if you saw the same friend again, you might make a different choice.

	%Previous work makes a strong case for both deterministic and random noise being relevant to behavior. For instance, deterministic, stimulus-driven noise is thought to be a much greater source of choice variability in perceptual decisions than random noise \citep{eeBrunton13}. Conversely random, neural noise is thought to drive exploratory singing behavior in song birds and the generation and control of this random noise has been linked to specific neural structures \citep{songbird2}.

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by deterministic processing of the stimulus \added[id=bob]{(which we refer to as `deterministic noise') vs other processes, including deterministic processing of unobserved stimuli as well as truly stochastic processes (which we refer to as `random noise')}. \added[id=siyu]{To distinguish between these two types of noise} we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus \added[id=bob]{(i.e., decision noise is purely deterministic noise), then people's choices should be identical for both decisions, since the stimulus is the same both times. Conversely, if the decision is a purely random function of the stimulus (i.e., decision noise is purely random noise), then people's choices will be different 50$\%$ of the time, since the random noise is different each time. In between these two extremes of purely deterministic and purely random drivers of behavioral variability, the extent to which people's decisions are consistent between the two decisions can be used to estimate the amount of deterministic and random noise.} %\added[id=siyu]{We built a computational model to quantify the relative magnitudes of stimulus-driven deterministic noise and non-stimulus-driven random noise in driving random exploration. Our model could detect the existence of previously unobserved deterministic processes that arise from the stimuli without explicitly knowing what the deterministic process is.}

    \added[id=bob]{In the following, we analyze behavior on the repeated decisions version of the Horizon Task in both a model-free and model-based manner. Our model-free analysis estimates the extent to which people's behavior is consistent across repeated versions of the same decision. By measuring how this choice consistency changes as a function of horizon, this model-free analysis offers qualitative insight into the extent to which behavioral variability is driven by deterministic vs random noise. Our model-based analysis uses a computational model of the explore-exploit decision in the Horizon Task that incorporates both noise processes. By fitting this model to the behavioral data, this model-based analysis allows us to quantify the relative size of the two sources of noise and how they change in the service of exploration.}

    % to the behavioral data.

    % of behavior that incorporates the two noise processes.

    % Our model-based analysis uses a

    % Thus, by analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from deterministic processing of the stimulus. This does not prove that random exploration is entirely deterministic or stochastic, but provides a lower bound on how much deterministic processes contribute to random exploration\added[id=siyu]{, and an upper bound on how much true random noises contribute to random exploration. Our work sheds light on how random exploration may be partially implemented by amplifying the effects of task-irrelevant stimuli on choice.} \deleted[id=siyu]{This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.}

	% SIYU says: "Conversely, the more their decision is driven by other processes, including both stochastic and unobserved deterministic processes, the less consistent their behavior should be. " is not correct. Unobserved deterministic processes will be the same for repeated games and will not contribute to choice inconsistency.

@@

	\section*{Results}

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' \replaced[id=siyu]{in explore-exploit}{on people's} decisions (Figure \ref{fig:taskfig}). In this task, participants make a series of \added[id=bob]{BE CAREFUL USING THE WORD REPEATED. ONLY USE IT TO REFER TO THE REPEATED GAMES - HERE I REPLACED REPEATED WITH `SERIES OF'} choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

    \added[id=bob]{THERE NEEDED TO BE MORE INFORMATION ABOUT THE HORIZON TASK HERE - REMEMBER THIS COUDL BE THE FIRST TIME PEOPLE HAVE SEEN THE TASK. The Horizon Task has two key features that together allow it to quantify explore-exploit behavior. The first of these features is  the time horizon --- the number of decisions participants will make in the future. By changing this horizon from short (1 trial) to long (6 trials), the Horizon Task allows us to control the relative value of exploration and exploitation. Just like the restaurant example in the introduction, when the horizon is short, participants should be more likely to exploit the option they believe to be best, because this leads to the highest payoff in the short term. Conversely, when the horizon is long, participants should be more likely to explore at first, because this allows them to gather information to make better choices later on. By contrasting behavior between short and long horizon conditions {\it on the very first trial}, when all else is equal, the Horizon Task allows us to quantify how behavior changes when it is valuable to explore.

    The second key feature of the Horizon Task are the four `example plays' at the start of each game that allow us to control exactly what participants know about the two bandits before they make their choice. In these example plays, participants are instructed which of the bandits to play allowing us to control how much information they have about each of the options. The instructed trials are used to set up one of two information conditions and `unequal information' or [1 3] condition, in which participants play one option once and the other three times, and an `equal information' or [2 2] condition, in which participants play both options twice.}

	% Crucially, before people make their first choice in the Horizon Task, they are given information about the mean payoff from each bandit in the form of four example plays distributed either unequally between bandits (i.e. 1 play of one bandit, 3 plays of the other, the [1 3] condition) or equally (2 plays each, the [2 2] condition). These example plays allow us to manipulate exactly what people know about each option before they make their first choice.

	Relative to the original Horizon Task, the key modification in this paper is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games separated by several minutes in time so as to avoid detection. By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. \added[id=bob]{Thus, if their behavior is a deterministic function of the stimuli, then they will make the same decision in both games and their choices will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games will be inconsistent some fraction of the time.} \replaced[id=siyu]{The extent to which participants' choices are consistent on the repeated versions of the games allow}{On average participants played 67.31 repeated games each allowing} us to quantify the extent to which \added[id=siyu]{the variability in} their behavior was \replaced[id=siyu]{driven by a deterministic process vs a random noise process}{a deterministic function of the stimulus or not}.

	%, the example plays allow us to probe how participants respond to the exact same explore-exploit choice twice.

	%These `repeated games' are the key manipulation in this paper and allow us to distinguish between deterministic and random sources of noise.  Specifically, if noise is deterministicly driven, then choices on repeated games should be consistent. Conversely if noise is randomly driven, then choices on repeated games should be inconsistent.

	\subsection*{Both behavioral variability and information seeking increase with horizon}

    \added[id=bob]{GENERAL COMMENT - WE NEED TO WALK PEOPLE THROUGH THE BASIC RESULTS MUCH MORE SLOWLY, OTHERWISE WILL BE CONFUSING FOR PEOPLE NEW TO THE HORIZON TASK.}

	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. %As in our previous work, we find evidence for two types of exploration in the Horizon Task: random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information. 
    \added[id=bob]{In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future.} %(Figure \ref{fig:modelfree}, Supplementary Figure S1).
    % As in our previous work, we find evidence for two types of exploration in the Horizon Task: random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information (Figure \ref{fig:modelfree}, Supplementary Figure S1).

    \subsubsection*{Model-free analysis}
    \added[id=bob]{
    In the model-free analysis, we quantify random and directed exploration using simple choice probabilities. Random exploration is quantified as the probability of choosing the option that has the lower average payout in the example plays in the equal, or [2 2], condition, $p(\mbox{low mean})$. The idea here is that, in the equal condition, the optimal strategy is to compute the mean payout for each bandit from the example plays and then always choose the option with the highest mean. When participants do not choose the option with the higher mean, the assumption is that this is due to some kind of `decision noise,' making the probability of choosing the low mean option a measure of behavioral variability. In this view, random exploration corresponds to an increase in $p(\mbox{low mean})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}A; t(64) = 7.99, p $<$ 0.001 for [2 2]).}

    \added[id=bob]{
    Directed exploration is quantified as the probability of choosing the more informative option $p(\mbox{high info})$ in the unequal, or [1 3], condition. The more informative option is the option played once during the example plays as choosing this option gives relatively more information (doubling the number of samples from 1 to 2) than choosing the option played three times (only increasing the number of sample by a third, from 3 to 4). In this view, directed exporation corresponds to an increase in $p(\mbox{high info})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}B; t(64) = 6.92, p $<$ 0.001).}

    \subsubsection*{Model-based analysis}

    \added[id=bob]{I'M ADDING THIS SECTION, BECAUSE IT ALLOWS US TO INTRODUCE THE MODEL EARLY. THAT WAY WHEN THE MODEL COMES IN LATER WE CAN SEE HOW IT'S CHANGED}

    \added[id=bob]{Another approach to understanding behavior in the Horizon Task is to use a computational model. In this case, we model  participants' choices on the first free-choice trial by assuming they make decisions by computing the difference in value (or utility) $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
	\begin{equation}
    \label{eq:origmodel}
		\Delta Q= \Delta R+A \Delta I+b+n
	\end{equation}
    where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of rewards shown on the forced trials, and $\Delta I$, the difference in information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right, and in [2 2] condition, $\Delta I$ is 0. $n_{det}$ denotes the deterministic noise, which is identical on the repeat versions of each game; and $n$ denotes decision noise, which, in this version of the model is a combination of deterministic and random noise. $n$ is assumed to come from a logistic distributions with mean 0 and standard deviations $\sigma$. }

    \added[id=bob]{The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon and information condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C, D; STATS FOR CHANGE IN INFO BONUS AND DECISION NOISE). }

@@

	\subsubsection*{\replaced[id=siyu]{Model validation}{Parameter recovery}\label{ch:appendix:bayesrecovery}}
	To be sure that our fit parameter values were meaningful \added[id=siyu]{and to understand the limits of our model, we evaluated our model extensively using simulated data. \added[id=bob]{This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis are presented in the Supplementary Materials.

@@

    Thus, overall, we were able to detect both deterministic and random noises using our model. Our model provides a lower bound for deterministic noise and an upper bound for random noise. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
	}

>>>>

\title{Separating random and deterministic sources of computational noise in explore-exploit decisions}
\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 15$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 85$\%$ of the variability is truly `random,' although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

	Imagine trying to decide where to go to dinner on a date. You can go to your favorite restaurant, the one you both really enjoy and always go to, or you can try a new restaurant that you know nothing about. Such decisions, in which we must choose between a well-known `exploit' option and a lesser known `explore' option, are known as explore-exploit decisions.  From a theoretical perspective, making optimal explore-exploit choices, i.e. choices that maximize long-term reward, is computationally intractable in most cases \citep{eegittins74, eeBasu18}. In part because of this computational complexity, there is considerable interest in how humans and animals solve the explore-exploit dilemma in practice {\citep{Mehlhorn15, SCHULZ20197,WILSON202149}.
	%\citep{eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90}, also known as random exploration. In this strategy, high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. From a modeling perspective, random exploration works by adding `decision noise' to the value of the options such that sub-optimal exploratory options can sometimes have a higher total score (i.e., value + noise) than the exploit option and get chosen. Such random exploration, is surprisingly effective and, if implemented correctly, can come close to optimal performance \citep{eethompson33, eebridle90, eeAgrawal11, eeChapelle11}.

	It has recently been shown that humans appear to use random exploration and can increase  decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014, Findling19}. In one of these tasks, known as the Horizon Task \citep{wilson2014}, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you will be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	One limitation of this previous research, however, is that it is difficult to tell whether what we have called `decision noise' actually reflect a noise process. From a modeling perspective, decision noise as defined in previous research essentially quantifies the extent to which behavior  cannot be explained by a computational model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be random noise. For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take the ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

	% That is, whether behavioral variability is due to genuinely stochastic processes in the brain or whether it is due to deterministic processes that we failed to observe.

	% From a modeling perspective, behavioral variability is essentially the variance that can not be explained by a model and is modeled as the level of decision noise. However, what we have called "decision noise" in previous researches could actually just be missing deterministic components from the model, it is difficult to tell whether decision noise truly arises from a stochastic process. Here we show that, while both random and deterministic noise drive variability in behavior, the noise driving random exploration is predominantly random. This suggests that random exploration depends on adaptive noise processes in the brain which are subject to cognitive control.

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by deterministic processing of the stimulus (which we refer to as `deterministic noise') vs other processes, including deterministic processing of unobserved stimuli as well as truly stochastic processes (which we refer to as `random noise'). To distinguish between these two types of noise, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus (i.e., decision noise is purely deterministic noise), then people's choices should be identical for both decisions, since the stimulus is the same both times. Conversely, if the decision is a purely random function of the stimulus (i.e., decision noise is purely random noise), then people's choices will be different 50$\%$ of the time, since the random noise is different each time. In between these two extremes of purely deterministic and purely random drivers of behavioral variability, the extent to which people's decisions are consistent between the two decisions can be used to estimate the amount of deterministic and random noise.

	% Our model could detect the existence of previously unobserved deterministic processes that arise from the stimuli without explicitly knowing what the deterministic process is.

	In the following, we analyze behavior on the repeated decisions version of the Horizon Task in both a model-free and model-based manner. Our model-free analysis estimates the extent to which people's behavior is consistent across repeated versions of the same decision. By measuring how this choice consistency changes as a function of horizon, this model-free analysis offers qualitative insight into the extent to which behavioral variability is driven by deterministic vs random noise. Our model-based analysis uses a computational model of the explore-exploit decision in the Horizon Task that incorporates both noise processes. By fitting this model to the behavioral data, this model-based analysis allows us to quantify the relative size of the two sources of noise and how they change in the service of exploration.

@@

	\section*{Results}	
	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' in explore-exploit decisions (Figure \ref{fig:taskfig}). In this task, participants make a series of choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. They are asked to choose between the two bandits to maximize the total rewards. One bandit always has a higher mean payout than the other. Participants need to try each bandit a few times to learn about the distribution of payout from that bandit. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

	The task is organized in games (Figure \ref{fig:taskfig}A). The mean payout of the two bandits are held fixed within a game and reset between games. Each game consists of either 5 or 10 trials. In the first 4 trials, participants are instructed about which bandit to choose, this allows us to manipulate what information from both bandits participants receive before they make their first free choice between the two bandits. From the 5th trial, participants make free choices between the two bandits. Participants have either 1 or 6 free choices to make.

	The Horizon Task has two key features that together allow it to quantify explore-exploit behavior. The first of these features is  the time horizon --- the number of decisions participants will make in the future. By changing this horizon from short (1 free-choice trial) to long (6 free-choice trials), the Horizon Task allows us to control the relative value of exploration and exploitation. Just like the restaurant example in the introduction, when the horizon is short, participants should be more likely to exploit the option they believe to be best, because this leads to the highest payoff in the short term. Conversely, when the horizon is long, participants should be more likely to explore at first, because this allows them to gather information to make better choices later on. By contrasting behavior between short and long horizon conditions {\it on the very first free-choice trial}, when all else is equal, the Horizon Task allows us to quantify how behavior changes, when it is more valuable to explore.

	The second key feature of the Horizon Task are the 4 `forced-choice trials' at the start of each game that allow us to control exactly what participants know about the two bandits before they make their choice. In these forced-choice trials, participants are instructed which of the bandits to play allowing us to control how much information they have about each of the options. The forced-choice trials are used to set up one of two information conditions: an `unequal information' or [1 3] condition, in which participants play one bandit once and the other three times, and an `equal information' or [2 2] condition, in which participants play both bandits twice.

	Relative to the original Horizon Task, the key modification in this paper is to give people `repeated games' (Figure \ref{fig:taskfig}B), in which they see exact same set of forced-choice plays twice in two separate games separated by several minutes in time so as to avoid detection. By repeating the forced-choice plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, then they will make the same decision in both games and their choices will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games will be inconsistent some fraction of the time. The extent to which participants' choices are consistent on the repeated versions of the games allow us to quantify the extent to which the variability in their behavior was driven by a deterministic process vs a random noise process.

	\subsection*{Both behavioral variability and information seeking increase with horizon}
	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results using both a model-free and model-based approach \citep{wilson2014}. In both analyses, we focus on just the first free-choice trial in each game, where the only thing that differs between the horizon conditions is the number of choices that participants will make in the future.

	\subsubsection*{Model-free analysis}
	In the model-free analysis, we quantify random and directed exploration using simple choice probabilities. Random exploration is quantified as the probability of choosing the option that has the lower average payout in the forced-choice plays in the equal, or [2 2], condition, $p(\mbox{low mean})$. The idea here is that, in the equal condition, the optimal strategy is to compute the mean payout for each bandit from the example plays and then always choose the option with the highest mean. When participants do not choose the option with the higher mean, the assumption is that this is due to some kind of `decision noise', making the probability of choosing the low mean option a measure of behavioral variability. In this view, random exploration corresponds to an increase in $p(\mbox{low mean})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}A; t(64) = 7.99, p $<$ 0.001).

	Directed exploration is quantified as the probability of choosing the more informative option $p(\mbox{high info})$ in the unequal, or [1 3], condition. The more informative option is the option played once during the forced-choice plays as choosing this option gives relatively more information (doubling the number of samples from 1 to 2) than choosing the option played three times (only increasing the number of sample by a third, from 3 to 4). In this view, directed exporation corresponds to an increase in $p(\mbox{high info})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}B; t(64) = 6.92, p $<$ 0.001).

	\subsubsection*{Model-based analysis}	
	Another approach to understanding behavior in the Horizon Task is to use a computational model. In this case, we model  participants' choices on the first free-choice trial by assuming they make decisions by computing the difference in value (or utility) $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
	\begin{equation}
		\label{eq:origmodel}
		\Delta Q= \Delta R+A \Delta I+b+n
	\end{equation}
	where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of rewards shown on the forced-choice trials, and $\Delta I$, the difference in information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right in the [1 3] condition, and in [2 2] condition, $\Delta I$ is 0.

	Here, $n$ denotes decision noise, which, in this version of the model is a combination of deterministic and random noise. $n$ is assumed to come from a logistic distributions with mean 0 and standard deviations $\sigma$.

	The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon and information condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C, D; STATS FOR CHANGE IN INFO BONUS AND DECISION NOISE).

@@

\subsubsection*{\replaced[id=siyu]{Model validation}{Parameter recovery}\label{ch:appendix:bayesrecovery}}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. \added[id=bob]{This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis are presented in the Supplementary Materials. }

@@

	Thus, overall, we were able to detect both deterministic and random noises using our model. Our model provides a lower bound for deterministic noise and an upper bound for random noise. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games.

MOVE THE REST OF THIS SECTION TO THE SUPPLEMENTARY MATERIAL
Firstly, we checked if our fitted deterministic noise could indeed capture unobserved deterministic process that was not accounted for by the decision model. We test this by leaving out one known deterministic process from the decision model, and ask if our method could recover that known deterministic process as deterministic noise. In particular, we fit a reduced version of our model that only considers reward and ignores the influence of uncertainty condition on explore-exploit decisions.  
$$\Delta Q= \Delta R+n_{det}+n_{ran}$$
Here, $\Delta Q, \Delta R, n_{det}, n_{ran}$ represent the same variables as in the full model.	If deterministic noise in our model can indeed capture unobserved deterministic processes that's missed by the model, then we would expect to see a higher level of fitted deterministic noise in the reduced model compared to in the full model, whereas the level of random noise should remain unchanged. By comparing the	fitted posterior distributions over the group-level means of the deterministic and random noise parameters $\sigma_{det}$ and $\sigma_{ran}$, as expected, we observed an increase in deterministic noise and no change in random noise between the reduced and the full model (Figure \ref{fig:reducedmodel}). This suggests that our model is capable of detecting missing deterministic processes.
